Simulating High-Dimensional Multivariate Data

### 2021-03-01 010-background.Rmd
Finish revising prior to passing to Alex

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in a practical timescale.

>>>>

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.

### 2021-03-01 index.Rmd
Finish revising prior to passing to Alex

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19]. For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

@@

Another central issue lies in characterizing dependency between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
We do, however, provide Pearson matching with some caveats.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure. Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm. The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.

>>>>

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19]. For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

@@

Another central issue lies in characterizing dependence between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
We do, however, provide Pearson matching with some caveats.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure. Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm. The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.
